---
title: "Spatial Modelling with SEM"
author: "G1"
date: "2025-03-09"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
    df_print: paged
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. SPATIAL

## Libraries

```{r}

library(sf)
library(ggplot2)
library(dplyr)
library(spdep)
library(spatialreg)
library(car)



```

## Read dataset

```{r cars}


nrb <- read.csv("/Users/stephen/MEGA/1. SCHOOL/3. Junior/1st Sem/STA3010 STATISTICAL MODELING/spartial/Combined_Data.csv")
#View
head(nrb)


```

## Dataset summary and conversion

```{r}

str(nrb)
max(nrb$x_5)
unique(nrb$x_6)
unique(nrb$x_8)
unique(nrb$x_9)


# x8, x9 categorical
# int var are x_5, x_7, X

ordlvl <- c("Very Low", "Low", "Medium", "High", "Very High")


nrb$y_11 <- as.numeric(factor(nrb$y_11, levels = ordlvl))
nrb$y_12 <- as.numeric(factor(nrb$y_12, levels = ordlvl))
nrb$y_13 <- as.numeric(factor(nrb$y_13, levels = ordlvl))
nrb$y <- as.numeric(factor(nrb$y, levels = ordlvl))
nrb$x_1 <- as.numeric(nrb$X)
nrb$x_2 <- as.numeric(nrb$X)
nrb$x_3 <- as.numeric(nrb$X)
nrb$x_4 <- ifelse(nrb$x_4 == "Yes", 1, 0)
nrb$x_5 <- as.numeric(nrb$x_5)
nrb$x_6 <- ifelse(nrb$x_6 == "Physical", 1, ifelse(nrb$x_6 == "Online", 2, 3))
nrb$x_7 <- as.numeric(nrb$x_7)
nrb$x_8 <- as.numeric(nrb$x_8)
nrb$x_9 <- as.numeric(nrb$x_9)
```

### Variables

y_11:  Access to new technology (Access)

y_12 :Openness, reception, utilization of new technology (Openness)

y_13 :Coping with modern technology (Coping)

Y : Adoption of New Technology (Adoption)

x_1 : Competency in ICT (ICTComp)

x_2 : Willingness to continuous training education and qualification
(Trainedu)

x_3 : Mastering digital transformation (Mdigtech)

x_4 : Percentage of employees with new technology (Newtech)

x_5:Percentage of employees with digital technical skills
(Techemployees)

 x_6 : Percentage of SMEs operating online (Moperate)

x_8: Financial Institutions as main financiers (Fininstmain)

x_9: Finance is the main barrier (Finbarrier)   

## Converting to sf

```{r}

names(nrb)[names(nrb) == "lat"] <- "latitude"
names(nrb)[names(nrb) == "long"] <- "longitude"

nrb_sf <- st_as_sf(nrb, coords = c("longitude", "latitude"), crs = 4326)
head(nrb_sf)


```

## Loading shapefile

```{r}


kenya_lvl3 <- st_read("/Users/stephen/MEGA/1. SCHOOL/3. Junior/1st Sem/STA3010 STATISTICAL MODELING/spartial/gadm41_KEN_shp/gadm41_KEN_2.shp")

# Check available column names to identify the correct column for subcounties
print(names(kenya_lvl3))

# Filter to keep only Nairobi subcounties
nairobi_shp <- kenya_lvl3[kenya_lvl3$NAME_1 == "Nairobi", ]

# Verify the filter worked
print(unique(nairobi_shp$NAME_1))  # Should only show "Nairobi"

# Compute centroids for labeling
nairobi_shp$centroid <- st_centroid(nairobi_shp$geometry)
nairobi_shp$lon <- st_coordinates(nairobi_shp$centroid)[, 1]
nairobi_shp$lat <- st_coordinates(nairobi_shp$centroid)[, 2]

# Plot Nairobi Subcounties with Labels
ggplot(data = nairobi_shp) +
  geom_sf(fill = "lightblue", color = "black") +  # Subcounty boundaries
  geom_text(aes(x = lon, y = lat, label = NAME_2), size = 3, color = "red") +  # Subcounty labels
  ggtitle("Nairobi Subcounties with Names") +
  theme_minimal()


```

## Combining all Embakasi

```{r}

# Merge all "Embakasi" subcounties into one
nairobi_shp$NAME_2 <- ifelse(grepl("embakasi", nairobi_shp$NAME_2, ignore.case = TRUE), "Embakasi", nairobi_shp$NAME_2)

# Dissolve to merge polygons of the same subcounty
nairobi_shp_updated <- nairobi_shp %>%
  group_by(NAME_2) %>%
  summarize(geometry = st_union(geometry), .groups = "drop")

# Compute centroid for each subcounty (for labeling)
nairobi_centroids <- st_centroid(nairobi_shp_updated)

# Extract centroid coordinates
nairobi_centroids <- nairobi_centroids %>%
  mutate(lon = st_coordinates(.)[,1], lat = st_coordinates(.)[,2])

# Plot updated subcounty map with labels
ggplot(data = nairobi_shp_updated) +
  geom_sf(fill = "lightblue", color = "black") +
  geom_text(data = nairobi_centroids, aes(x = lon, y = lat, label = NAME_2), size = 4, color = "red") +  
  ggtitle("Updated Nairobi Subcounties (Merged Embakasi)") +
  theme_minimal()


```

## Merging shapefile and dataset

```{r}


# Ensure consistent naming for merging
nrb$ADM2_EN <- trimws(tolower(nrb$ADM2_EN))  # Convert dataset subcounty names
nairobi_shp_updated$NAME_2 <- trimws(tolower(nairobi_shp_updated$NAME_2))  # Convert shapefile subcounty names

# Merge the dataset with the shapefile
nrb_sf <- left_join(nairobi_shp_updated, nrb, by = c("NAME_2" = "ADM2_EN"))



```

## Check missing

```{r}


sum(is.na(nrb_sf))
sum(is.na(nrb))

nrb_sf[apply(is.na(nrb_sf), 1, any), ]
nrb_sf <- na.omit(nrb_sf)


```

# II. Spatial matrix creation

```{r}

# Create spatial weights matrix (Queen contiguity)
nb <- poly2nb(nrb_sf)  # Creates neighborhood list
W <- nb2listw(nb, style = "W")  # Converts to weights list
W

# Queen
plot(st_geometry(nrb_sf), main = "Queen Contiguity Neighbors")
plot(nb, st_coordinates(st_centroid(st_geometry(nrb_sf))), add = TRUE, col = "blue")

# Rook
nb_rook <- poly2nb(nrb_sf, queen = FALSE)  # Creates Rook contiguity neighborhood list
plot(st_geometry(nrb_sf), main = "Rook Contiguity Neighbors")
plot(nb_rook, st_coordinates(st_centroid(st_geometry(nrb_sf))), add = TRUE, col = "red")

# Distance
coords <- st_coordinates(st_centroid(st_geometry(nrb_sf)))
nb_dist <- dnearneigh(coords, 0, 2000)  # Neighbors within 2km

plot(st_geometry(nrb_sf), main = "Distance-Based Neighbors (2km)")
plot(nb_dist, coords, add = TRUE, col = "green")




```

# III. Moran test

```{r}


# Moran's I test for spatial autocorrelation
moran_result <- moran.test(nrb_sf$y, listw = W)
print(moran_result)

# Moran's I test for spatial autocorrelation
moran_result <- moran.test(nrb_sf$y_11, listw = W)
print(moran_result)

# Moran's I test for spatial autocorrelation
moran_result <- moran.test(nrb_sf$y_12, listw = W)
print(moran_result)

# Moran's I test for spatial autocorrelation
moran_result <- moran.test(nrb_sf$y_13, listw = W)
print(moran_result)



```

## Collinearity check

```{r}   vif(lm(y13 ~ x_1 + x_4 + x_5 + x_7 + x_8 + x_9, data = nrb_sf))}
```

## SLM model (SAR)

```{r}


# Fit a Spatial Lag Model (SAR)
sar_model <- lagsarlm(y_13 ~ x_1 + x_4 + x_5 + x_7 + x_8 + x_9, 
                      data = nrb_sf, listw = W)
summary(sar_model)


```

## SEM model

```{r}


# Fit a Spatial Error Model (SAE)
sae_model <- errorsarlm(y_13 ~ x_1 + x_4 + x_5 + x_7 + x_8 + x_9, 
                      data = nrb_sf, listw = W)
summary(sae_model)


```

## SDM model

```{r}

# Fit a Spatial Durbin Model (SDM)
sdm_model <- lagsarlm(y_13 ~ x_1 + x_4 + x_5 + x_7 + x_8 + x_9, 
                      data = nrb_sf, listw = W, type = "mixed")
summary(sdm_model)


```

# IV. Model selection

## Langrange multiplier test

```{r}

# Run LM tests on a regular linear model
lm.LMtests(lm(y_13 ~ x_1 + x_4 + x_5 + x_7 + x_8 + x_9, 
                      data = nrb_sf), 
           listw = W, test = "all")


```

## AIC comparison

```{r}

# Compare models using AIC
aic_values <- c(SAR = AIC(sar_model), 
                SAE = AIC(sae_model), 
                SDM = AIC(sdm_model))
sort(aic_values)


```
